Giant Low Surface Brightness Galaxies As Merger Remnants

### 2015-05-07 glsb.tex
blah

If GLSB disks are indeed the remains of accreted dwarf galaxies, they represent a unique opportunity to observationally probe the outer regions of galaxy dark matter halos, especailly if they form around elliptical galaxies.  The mass profiles of elliptical galaxies have been difficult to constrain since they lack the HI disks found around spirals.

@@

In this paper, we use IFU spectroscopy to measure the star formation histories of two GLSB systems.

The problem with considering GLSB systems as normal spirals is that they are clearly outliers when one considers the specific angular momentum of the systems.  The difference between high surface brightness (HSB) spirals and low surface brightness spirals (LSB), could simply be an issue of specific angular momentum. xxx-this goes in the discussion I think. \citet{Dutton12} is a nice paper on how disk galaxies get their angular momentum--I'll need to make a calc of how much J is in the GLSB disks to compare it.

GLSBs are typically classified as spiral galaxies \citep{Das13}, however, high resolution imaging of Malin~1 has shown that the central region appears to be a normal

\citet{Raskutti14} use VIRUS-P to look at stellar halo kinematics of massive ellipticals.  Lots of good stuff on fast/slow rotators.  Also a very nice plot showing the radial gradients in age and metallicity as measured from Lick indices.  They don't see any sharp transition like I do, all very smooth age and metal curves.

@@

\citet{Pickering97} has the \HI\ rotation curves of these guys.  For UGC 6614, has M \HI\ $= 2.5 \pm0.2\times 10^{10} \msun h_{75}^{-2}$.  And $3.6 \pm 0.4 \times 10^{10} \msun h_{75}^{-2}$for Malin 2.  Notes that these are galaxies that are massive and DM dominated.  In Malin 2, they find some gas moving relative to the regular rotating disk, and suspect that a dwarf fell though and stimulated star formation and this could be how stars managed to form in the low density environment.

Can we say that the stars formed in a higher density environment and then got dispersed.

\citet{Das13} look at the environments of GLSB systems.  He says that GLSBs are an extreme form of late type spiral galaxies.  I think I can show that no, they are really normal elliptical galaxies that happen to have accreted disks.

Nice paper on galaxies that live in voids and how one did manage to build a polar \HI\ disk with no stars: \citet{Stanonik09}

Chemical abundances in a polar disk:  \citet{Spavone10}

a near scoop by \citet{Lelli10}.  They re-analyze \HI\ data on Malin 1 and one other.  But a nice place to look for introduction material.  They point out that GLSB galaxies might be normal HSB galaxies in the center.  They claim to see steeply rising rotation curves, and compare to formation mechanisms.

\citet{Morganti97} look at an elliptical with an \HI\ disk.  Find a nice flat rotation curve.
\citet{Emonts08}  Another massive \HI\ ring around an S0 galaxy.
\citet{Weijmans08} another \HI\ ring around an early type.  Anne-Marie has lots of SAURON papers pushing out to large radii in ellipticals.

\citet{Pen06} look at forming extended disks by disrupting a dwarf galaxy.

@@

\cite{Rahman2007} has Spitzer observations of a few GLSB galaxies.

\citet{Das10} find CO in Malin 2.  Make some points about it being stable

\citet{Mapelli12} making ring galaxies from off center collisions.

\citet{Mapelli08} look at if ring galaxies could evolve into GLSB systems.

@@

\citet{Schweizer89} Do a similar study with IC 2006, where they observe an \HI\ ring around the elliptical galaxy to get the total mass at large radii.  Looks like they get out to about 12.5 kpc.

\citet{Oosterloo07} HIPASS detection of \HI\ gas around 30 early-type galaxies.  Most have regular disk or ring kinematics.  Suggest these are ancient.  They don't seem to have the resolution to do in-depth kinematics and DM halo stuff.

\citet{Magain13} look at strong lenses around 25 ellipticals, and then claim they see no evidence for extended DM halos.  Confusing plot--Maybe they get out to 5 $R_{1/2}$?  They say they only need a factor of 2, that could come from baryonic dark matter.

\citet{Spray95} CCD imaging of 8 GLSB systems.

\citet{Morganti13} Do modeling of ellipticals with declining velocity dispersion profiles.  Good place to start looking for the best way to model the long-slit data.

\citet{Buote12} review of dark matter in ellipticals from x-ray observations.

XXX-need to look up some Thilker papers on extended UV disks, since these objects could be similar.  From his talk, UGC 1382 shows up in stripe 82 coadd (and NUV).  UGC 3642, GALEX shows a disk, and \HI\ is counter rotating.

\citet{Salim10} Looks at HST imaging of z = 0.1 early type galaxies that have UV excess, and finds some extended disks.  They think the gas and star formation is coming from minor mergers or accretion.

\citet{Salim12} Finds 2 more GLSB type objects -- could be a sign that there are multiple routes to making GLSB systems.

@@

The other way to measure the dark matter halo of an elliptical is to use gravitational lensing.  \citet{Belokurov07} find a large ring in SDSS.

@@

\begin{figure}
\epsscale{.4}
\plotone{Plots/Color_Magnitu_Diagram.eps}\plotone{Plots/Faber_Jackson_Relation.eps}\\
\plotone{Plots/Fundamental_Plane.eps}\plotone{Plots/mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin 2 and UGC 6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band Fundamental Plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   Looking at the SDSS data, our GLSB galaxies appear as normal red elliptical galaxies.  \label{elliptical_relations}}
\end{figure}

@@

%/astro/store/shared-scratch1/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, Malin 1 from \citet{Moore06}, Malin 2 and UGC 6614 from \citet{Pickering97}.  Here, $R_{max}$ is the maximum radius with an \HI\ detection for the THINGS galaxies and the maximum radius with a stellar flux detection for the GLSB galaxies.  It is interesting to note that the stellar disks of GLSB galaxies are often larger than the \HI\ disks of nearby systems.  \label{disk_comp}}
\end{figure}

>>>>

If GLSB disks are indeed the remains of accreted dwarf galaxies, they represent a unique opportunity to observationally probe the outer regions of galaxy dark matter halos, especailly if they form around elliptical galaxies.  The mass profiles of elliptical galaxies have been difficult to constrain since they lack the extended \HI\ disks found around spirals.  As an example of how difficult it can be to measure the outer regions of ellipical galaxy kinematics, \citet{Murphy2011} model M87 out to $\sim$23 kpc with stellar kinematics and out to 51 kpc with globular clusters (or $R=4.8 R_e$, where $R_e$ is the effective radius of M87). \citet{Raskutti14} are also limited to measuring kinematics out to $\sim$2-5 $R_e$.

In comparison, the stellar disk of Malin~2 extends beyond XXX (XXX $R_e$), and the HI disk reaches beyond 100 kpc (XXX $R_e$).  These distances have occasionally been reached by x-ray observations \citep{Buote12}.

@@

%In this paper, we use IFU spectroscopy to measure the star formation histories of two GLSB systems.

%The problem with considering GLSB systems as normal spirals is that they are clearly outliers when one considers the specific angular momentum of the systems.  The difference between high surface brightness (HSB) spirals and low surface brightness spirals (LSB), could simply be an issue of specific angular momentum. xxx-this goes in the discussion I think. \citet{Dutton12} is a nice paper on how disk galaxies get their angular momentum--I'll need to make a calc of how much J is in the GLSB disks to compare it.

%\citet{Raskutti14} use VIRUS-P to look at stellar halo kinematics of massive ellipticals.  Lots of good stuff on fast/slow rotators.  Also a very nice plot showing the radial gradients in age and metallicity as measured from Lick indices.  They don't see any sharp transition like I do, all very smooth age and metal curves.

@@

%\citet{Pickering97} has the \HI\ rotation curves of these guys.  For UGC 6614, has M \HI\ $= 2.5 \pm0.2\times 10^{10} \msun h_{75}^{-2}$.  And $3.6 \pm 0.4 \times 10^{10} \msun h_{75}^{-2}$for Malin 2.  Notes that these are galaxies that are massive and DM dominated.  In Malin 2, they find some gas moving relative to the regular rotating disk, and suspect that a dwarf fell though and stimulated star formation and this could be how stars managed to form in the low density environment.

%Can we say that the stars formed in a higher density environment and then got dispersed.

%\citet{Das13} look at the environments of GLSB systems.  He says that GLSBs are an extreme form of late type spiral galaxies.  I think I can show that no, they are really normal elliptical galaxies that happen to have accreted disks.

%Nice paper on galaxies that live in voids and how one did manage to build a polar \HI\ disk with no stars: \citet{Stanonik09}

%Chemical abundances in a polar disk:  \citet{Spavone10}

%a near scoop by \citet{Lelli10}.  They re-analyze \HI\ data on Malin 1 and one other.  But a nice place to look for introduction material.  They point out that GLSB galaxies might be normal HSB galaxies in the center.  They claim to see steeply rising rotation curves, and compare to formation mechanisms.

%\citet{Morganti97} look at an elliptical with an \HI\ disk.  Find a nice flat rotation curve.
%\citet{Emonts08}  Another massive \HI\ ring around an S0 galaxy.
%\citet{Weijmans08} another \HI\ ring around an early type.  Anne-Marie has lots of SAURON papers pushing out to large radii in ellipticals.

@@

%\cite{Rahman2007} has Spitzer observations of a few GLSB galaxies.

%\citet{Das10} find CO in Malin 2.  Make some points about it being stable

%\citet{Mapelli12} making ring galaxies from off center collisions.  
%\citet{Mapelli08} look at if ring galaxies could evolve into GLSB systems.

@@

%\citet{Schweizer89} Do a similar study with IC 2006, where they observe an \HI\ ring around the elliptical galaxy to get the total mass at large radii.  Looks like they get out to about 12.5 kpc.

%\citet{Oosterloo07} HIPASS detection of \HI\ gas around 30 early-type galaxies.  Most have regular disk or ring kinematics.  Suggest these are ancient.  They don't seem to have the resolution to do in-depth kinematics and DM halo stuff.

%\citet{Magain13} look at strong lenses around 25 ellipticals, and then claim they see no evidence for extended DM halos.  Confusing plot--Maybe they get out to 5 $R_{1/2}$?  They say they only need a factor of 2, that could come from baryonic dark matter.

%\citet{Spray95} CCD imaging of 8 GLSB systems.

%\citet{Morganti13} Do modeling of ellipticals with declining velocity dispersion profiles.  Good place to start looking for the best way to model the long-slit data.

%\citet{Buote12} review of dark matter in ellipticals from x-ray observations.

%XXX-need to look up some Thilker papers on extended UV disks, since these objects could be similar.  From his talk, UGC 1382 shows up in stripe 82 coadd (and NUV).  UGC 3642, GALEX shows a disk, and \HI\ is counter rotating.

%\citet{Salim10} Looks at HST imaging of z = 0.1 early type galaxies that have UV excess, and finds some extended disks.  They think the gas and star formation is coming from minor mergers or accretion.

%\citet{Salim12} Finds 2 more GLSB type objects -- could be a sign that there are multiple routes to making GLSB systems.

@@

%The other way to measure the dark matter halo of an elliptical is to use gravitational lensing.  \citet{Belokurov07} find a large ring in SDSS.

@@

\begin{figure}
\epsscale{.4}
\plotone{Plots/Color_Magnitu_Diagram.eps}\plotone{Plots/Faber_Jackson_Relation.eps}\\
\plotone{Plots/Fundamental_Plane.eps}\plotone{Plots/mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin 2 and UGC 6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band Fundamental Plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   Looking at the SDSS data, our GLSB galaxies appear as realtively typical red elliptical galaxies.  \label{elliptical_relations}}
\end{figure}

@@

%/astro/store/shared-scratch1/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, Malin 1 from \citet{Moore06}, Malin 2 and UGC 6614 from \citet{Pickering97}.  Here, $R_{max}$ is the maximum radius with an \HI\ detection for the THINGS galaxies and the maximum radius with a stellar flux detection for Malin~1.  It is interesting to note that the \emph{stellar} disks of GLSB galaxies can be larger than the \HI\ disks of nearby systems.  \label{disk_comp}}
\end{figure}
